---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the data and convert the date strings into POSIX dates:
```{r}
data <- read.csv("activity.csv", header=TRUE, 
                 colClasses=c("numeric", "character", "numeric"),
                 stringsAsFactors=FALSE)
data$date <- as.Date(data$date, format="%Y-%m-%d")
```


## What is mean total number of steps taken per day?

Now make a histogram of the number of steps.  Here
we will ignore any missing values.

```{r}
library(ggplot2)
g <- ggplot(data, aes(steps))
g + geom_histogram(binwidth=15, fill='blue') + 
  labs(title="Fitness tracker activity", x="Number of Steps")
```

Find the mean/median number of steps and format them.
```{r histogram, fig.width=3, fig.height=2}
mn <- mean(data$steps, na.rm=TRUE)
md <- median(data$steps, na.rm=TRUE)
paste("The mean number of steps is", format(mn, digits=4),
      "and the median is", format(md, digits=2))
```

This is a rather skewed distribution with a large number
of zero values; perhaps some test subjects aren't actually
using their devices.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
