---
title: 'Fuel Efficiency: Automatic vs. Manual Transmission'
output: pdf_document
---

## Executive Summary
The effects of transmission type on fuel efficiency are investigated
using the mtcars data set and linear models.  A good fit is found
using the vehicle weight, quarter mile time, and transmission type.
The conclusion is that manual transmissions
are significantly more efficient (approximately 14 +- 3 miles per gallon)
than automatic transmissions with all other variables held constant.
There is also evidence that the fuel efficiency of
manual transmission cars declines more rapidly with the weight of the car,
although not to the extent that it compensates for the transmission
type in the sample examined.

## Exploratory Analysis and Model Selection
As shown in the appendix, there is a difference in the mean fuel efficiency
between manual and automatic transmission vehicles.  However, since
the transmission type is correlated with other variables (like vehicle
weight) that we expect to affect fuel efficiency, more extensive
modelling is required.  Here a linear analysis is used to study this
question.

There are a relatively small number of variables in this data set, 
so it is reasonable to use full subset selection, which computes the best 
model (ranked by R^2) by trying all combinations of features.
```{r subset}
library(leaps)
data(mtcars)
mtcars$am <- factor(mtcars$am, labels=c("Automatic", "Manual"))
regfit.full = regsubsets(mpg ~ ., mtcars, nvmax=11)
summary(regfit.full)$rsq
```
R^2 increases up to about 5 features, then levels off.  The variables included
are: `r regfit.full$xnames[summary(regfit.full)$which[5,]]`, 
which includes the transmission type.

Next, consider the variance inflation factors to exclude variables from
this set that are too strongly related.
```{r vif}
library(car)
fit.five <- lm(mpg ~ disp + hp + wt + qsec + am, mtcars)
vif(fit.five)
```
Disp is extremely correlated with at least one other variable.
Removing that and repeating shows that hp has a large VIF.  Once
that is removed the remaining VIFs are all less than 5, and hence
not significant.
```{r vif2}
fit.three <- lm(mpg ~ wt + am + qsec, mtcars)
vif(fit.three)
```

Next, is there any support for cross-terms?
```{r anova}
fit1 <- lm(mpg ~ wt + am + qsec, mtcars)
fit2 <- lm(mpg ~ wt + am + qsec + wt:am, mtcars)
fit3 <- lm(mpg ~ wt + am + qsec + wt:am + wt:qsec, mtcars)
fit4 <- lm(mpg ~ wt + am + qsec + wt:am + wt:qsec + am:qsec, mtcars)
anova(fit1, fit2, fit3, fit4)
```
Only the first cross term (wt by transmission type) improves
the fit significantly, but the F test for that term suggests
that it should be included.

The next step is to examine the residuals of the adopted fit.
These are shown in the appendix, but suggest no problems.
The residuals do not show strong evidence for non-linearity, and
the quantiles are roughly Normal.  Therefore, the following
model is adopted:
```{r model}
summary(fit2)$coefficients
```

## Interpretation
Examining the cofficients of the adopted fit shows that all
the measured relations are significant (based on their Pr(>|t|) values)
and the resulting R^2 value is better than a linear fit to all variables
without cross terms.

The coefficients show that manual transmission vehicles have,
on average, 14 +- 3 miles per gallon better fuel efficiency if all other 
variables are held fixed.  However,
manual transmission cars seem to suffer from an increased weight
penalty than those with automatic transmissions of -4 +- 1 mpg
per extra 1000lbs weight.  Therefore, an automatic car should have 
approximately the same fuel efficiency as a manual car
with all other variables held fixed but with the latter 
`r format(14.079 / 4.141 * 1000, digits=2)` lbs heavier.
This is a large weight difference, however, corresponding to
nearly the full range across the sample.

There are some important caveats.  Most importantly, this data
is from 1974, and there is no reason to expect these relations to
apply to modern cars.  Second, the number of cars in the data
set is modest.  This prevents us from evaluating more complex models
effectively.  Physically, a linear model should work better in terms
of fuel consumption rather than the efficiency measure (mpg) studied here.

## Appendix
```{r boxplot}
library(ggplot2)
library(gridExtra, quietly=TRUE)

g1 <- ggplot(mtcars) + geom_boxplot(aes(am, mpg)) + 
      xlab("Transmission Type") + ylab("Fuel Efficiency [mpg]")
g2 <- ggplot(mtcars) + geom_boxplot(aes(am, wt)) + 
    xlab("Transmission Type") + ylab("Weight [1000 lbs]")
grid.arrange(g1, g2, ncol=2)
```

```{r fitdiagnostics}
par(mfrow=c(2,2))
plot(fit2)
```