---
title: "Damage and Casualties from US Weather Events"
output: html_document
---

## Synopsis
The consequences, personal and financial, of weather events were studied 
in the US over the period 2001-2011 using NOAA data.  We find that 
tornadoes cause the largest amount of personal injury, while droughts
and floods cause the most financial damage.  However, on a per-event
level, hurricanes are the most damaging, especially financially.  The
damage caused by hurricanes fluctuates substantially from year to year
due to their rarity and where they happen to hit.  Thus, it may be
necessary to keep a large reserve of resources to effectivley respond
to hurricanes.

##Data Processing

It is first necessary to download the data if not present.  The dataset
consists of the NOAA weather event database, downloaded on 
2014/10/14 at 9:56pm UTC.

```{r getdata}
data.file <- "stormData.csv.bz2"
if (!file.exists(data.file)) {
    cat("Downloading data file")
    data.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(data.url, destfile=data.file, method="curl")
}
```

### Cleaning the data

The next step is to load it in and clean it up a bit.  
This is not a small file (902297 rows), 
so this may take a little while.  We remove columns not of interest for
the current analysis to save memory, and convert the beginning date to
a real date.

Both crop and property damage come with a metric specifier ("M", "K", "B" for
millions, thousands, billions, respectively).  In addition, there
are some mistakes for the exponent (H, 0, +, ?, etc.)  We remove these
rows because it isn't clear how to convert them to real values, and
multiply in the metric scale factor for the others.  The event
type is converted to a factor for future convenience, and 
the columns given nicer names.

This data set spans a long time period (1950-2011).  Because of the increase
in population density over this period (which significantly affects all
aspects of storm damage), this analysis is restricted to Jan 1st, 2001
and later.  Note that there are no incomplete cases in this dataset.
We also combine the Event types HURRICANE/TYPHOON and HURRICANE.
```{r readdata, cache=TRUE}
storm.data <- read.csv(bzfile(data.file), stringsAsFactors=FALSE,
                       nrows=903000)
# Do this cut first so we can count identified exponent cut losses
storm.data$BGN_DATE <- as.Date(storm.data$BGN_DATE, format="%m/%d/%Y")
desired.cols <- c("BGN_DATE", "FATALITIES", "INJURIES", "EVTYPE",
                 "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")     
storm.data <- subset(storm.data, BGN_DATE >= "2001/01/01", desired.cols)
nrows.storm <- nrow(storm.data)
storm.data$PROPDMGEXP <- toupper(storm.data$PROPDMGEXP)
storm.data$CROPDMGEXP <- toupper(storm.data$CROPDMGEXP)
good.exponents <- c("", "K", "M", "B")
storm.data <- subset(storm.data, PROPDMGEXP %in% good.exponents &
                    CROPDMGEXP %in% good.exponents)
cat(paste("Requiring understood exponents removed",
          nrows.storm - nrow(storm.data), "of", nrows.storm, "rows"))
storm.data$EVTYPE <- as.factor(storm.data$EVTYPE)
names(storm.data) <- c("Begin.Date", "Fatalities", "Injuries",
                        "Event.Type", "Property.Damage", 
                        "Property.Damage.Exp",
                        "Crop.Damage", "Crop.Damage.Exp")
cat(paste("Number of incomplete cases:", sum(!complete.cases(storm.data))))
```

Now multiply in the damage exponent scalings (e.g., Billions).
```{r damagescale, cache=TRUE}
val.prefix <- function(dmg, exponent) {
    if (exponent == "K") {dmg * 1e3}
    else if (exponent == "M") {dmg * 1e6}
    else if (exponent == "B") {dmg * 1e9}
    else dmg
}
storm.data$Property.Damage <- mapply(val.prefix, storm.data$Property.Damage, 
                                    storm.data$Property.Damage.Exp)
storm.data$Crop.Damage <- mapply(val.prefix, storm.data$Crop.Damage,
                                 storm.data$Crop.Damage.Exp)  
storm.data$Property.Damage.Exp <- NULL  # No longer needed
storm.data$Crop.Damage.Exp <- NULL
```

It is convenient for later to combine HURRICANE/TYPHOON events and
HURRICANE events, since the governmental response is similar.
```{r hurtyph}
whtyph <- which(storm.data$Event.Type=="HURRICANE/TYPHOON")
storm.data[whtyph, "Event.Type"] <- "HURRICANE"
```

### Inflation correction

Inflation is not included in the damage values.  This is not
a huge effect between 2001 and 2011, but is worth including because
the relative rarity of some major effects means that not correction
will introduce significant noise in the estimates.  Inflation can
be computed using information on the consumer price index, here
dowloaded on 2014/10/14 at 11:13pm UTC.  

```{r inflationdata}
cpi.file <- "CPI.csv"
if (!file.exists(cpi.file)) {
    cpi.url <- "http://research.stlouisfed.org/fred2/data/CPIAUCSL.csv"
    download.file(cpi.url, destfile=cpi.file, method="curl")
}
```

Convert this to a function that gives the CPI relative to
2001/01/01 using linear interpolation, and apply it to the damages.
This turns out to be *much* faster if we convert the dates to numeric:
```{r cpifun}
monthly.cpi <- read.csv(cpi.file, header=TRUE, stringsAsFactors=FALSE)
monthly.cpi$DATE <- as.Date(monthly.cpi$DATE, format="%Y-%m-%d")
monthly.cpi <- monthly.cpi[monthly.cpi$DATE > "2000-01-01",]
cpi.start <- monthly.cpi[monthly.cpi$DATE == "2001-01-01", "VALUE"]
cpi.interp <- approxfun(as.numeric(monthly.cpi$DATE), 
                        monthly.cpi$VALUE / cpi.start)
storm.data$rel.cpi <- sapply(as.numeric(storm.data$Begin.Date), cpi.interp)
storm.data$Property.Damage <- storm.data$Property.Damage * storm.data$rel.cpi
storm.data$Crop.Damage <- storm.data$Crop.Damage * storm.data$rel.cpi
```
Now all of the damages are in Jan 2001 dollars.

## Results

There are two ways of looking at the damages and injuries:
the mean consequences per event of type X and the total damage
over all events of type X.  

### Total damage by event type

Start by forming the summary statistics:
```{r makesummary}
library(plyr, quiet=TRUE)
storm.agg <- ddply(storm.data, ~ Event.Type, summarize,
                    mean.Fatalities=mean(Fatalities),
                    total.Fatalities=sum(Fatalities),
                    mean.Injuries=mean(Injuries),
                    total.Injuries=sum(Injuries),
                    mean.Crop=mean(Crop.Damage),
                    total.Crop=sum(Crop.Damage),
                    mean.Property=mean(Property.Damage),
                    total.Property=sum(Property.Damage))
```

There are too many types of events to easily visualize,
so here only the top 5 are shown (i.e., top fatalities, top injuries,
etc.).  Damages are converted to billions of dollars.

```{r toptotal}
nshow <- 5
top.Fatalities <- storm.agg[order(storm.agg$total.Fatalities, decreasing=T),
                            c("Event.Type", "total.Fatalities")]
top.Fatalities <- head(top.Fatalities, nshow)
top.Injuries <- storm.agg[order(storm.agg$total.Injuries, decreasing=T),
                            c("Event.Type", "total.Injuries")]
top.Injuries <- head(top.Injuries, nshow)
top.Property <- storm.agg[order(storm.agg$total.Property, decreasing=T),
                            c("Event.Type", "total.Property")]
top.Property <- head(top.Property, nshow)
top.Property <- mutate(top.Property, total.Property=total.Property*1e-9)
top.Crop <- storm.agg[order(storm.agg$total.Crop, decreasing=T),
                            c("Event.Type", "total.Crop")]
top.Crop <- head(top.Crop, nshow)
top.Crop <- mutate(top.Crop, total.Crop=total.Crop*1e-9)
```


```{r plotresults, fig.width=9, fig.height=5}
require(ggplot2) || stop("Install ggplot2")
require(grid) || stop("Install ggplot2")
fat.plot <- ggplot(top.Fatalities, aes(Event.Type, total.Fatalities)) +           
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Total Fatalities",x="Storm Type")
inj.plot <- ggplot(top.Injuries, aes(Event.Type, total.Injuries)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Total Injuries", x="Storm Type")
prop.plot <- ggplot(top.Property, aes(Event.Type, total.Property)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Total Property Damage [Billions of $]", 
                 x="Storm Type")
crop.plot <- ggplot(top.Crop, aes(Event.Type, total.Crop)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Total Crop Damage [Billions of $]", 
                 x="Storm Type")
grid.newpage()
vp <- viewport(layout = grid.layout(2, 4, height=unit(c(1, 6), "null")))
pushViewport(vp)
print(fat.plot, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
print(inj.plot, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
print(prop.plot, vp = viewport(layout.pos.row = 2, layout.pos.col = 3))
print(crop.plot, vp = viewport(layout.pos.row = 2, layout.pos.col = 4))

grid.text("Top 5 storm types in the US 2001-2011 by total damages", 
          vp = viewport(layout.pos.row = 1, layout.pos.col = 1:4))
```

As can be seen, Tornadoes cause the largest number of injuries and
fatalities, while floods and droughts cause the largest amount of
financial damage over the period studied.

### Mean damage per event

We can do the same thing by average damages per event.
Because of the way that tornadoes are
tracked in the NOAA database (when they cross county lines or 
temporarily lift off, they are counted as a new event) the mean
damage per event information is difficult to interpret for tornadoes.
Fixing this would be rather difficult, unfortunately, but it is 
worth keeping in mind.
```{r averagedmg, fig.width=9, fig.height=5}
mean.Fatalities <- storm.agg[order(storm.agg$mean.Fatalities, decreasing=T),
                            c("Event.Type", "mean.Fatalities")]
mean.Fatalities <- head(mean.Fatalities, nshow)
mean.Injuries <- storm.agg[order(storm.agg$mean.Injuries, decreasing=T),
                            c("Event.Type", "mean.Injuries")]
mean.Injuries <- head(mean.Injuries, nshow)
mean.Property <- storm.agg[order(storm.agg$mean.Property, decreasing=T),
                            c("Event.Type", "mean.Property")]
mean.Property <- head(mean.Property, nshow)
mean.Property <- mutate(mean.Property, mean.Property=mean.Property*1e-6)
mean.Crop <- storm.agg[order(storm.agg$mean.Crop, decreasing=T),
                            c("Event.Type", "mean.Crop")]
mean.Crop <- head(mean.Crop, nshow)
mean.Crop <- mutate(mean.Crop, mean.Property=mean.Crop*1e-6)
fat.mplot <- ggplot(mean.Fatalities, aes(Event.Type, mean.Fatalities)) +           
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Mean Fatalities",x="Storm Type")
inj.mplot <- ggplot(mean.Injuries, aes(Event.Type, mean.Injuries)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Mean Injuries", x="Storm Type")
prop.mplot <- ggplot(mean.Property, aes(Event.Type, mean.Property)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Mean Property Damage [Millions of $]", x="Storm Type")
crop.mplot <- ggplot(mean.Crop, aes(Event.Type, mean.Crop)) +
            geom_bar(stat="identity", fill="blue") +
            theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
            labs(y="Mean Crop Damage [Millions of $]", 
                 x="Storm Type")
grid.newpage()
vp <- viewport(layout = grid.layout(2, 4, height=unit(c(1, 6), "null")))
pushViewport(vp)
print(fat.mplot, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
print(inj.mplot, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
print(prop.mplot, vp = viewport(layout.pos.row = 2, layout.pos.col = 3))
print(crop.mplot, vp = viewport(layout.pos.row = 2, layout.pos.col = 4))

grid.text("Top 5 storm types in the US 2001-2011 by mean damages", 
          vp = viewport(layout.pos.row = 1, layout.pos.col = 1:4))
```

As can be seen, hurricanes cause far more damage (especially
financial damage) than the other types of events on a per-storm basis.
Because of the small number of hurricanes experienced in the US a year,
and the high cost associated with such events, the total damages
caused by Hurricanes every year will vary significantly, as
can be seen in the following plot:

```{r hurricanes, fig.width=5, fig.height=4}
library(lubridate, quiet=TRUE)
hurricane.data <- storm.data[storm.data$Event.Type == "HURRICANE",]
hurricane.data <- mutate(hurricane.data, 
                  Year = as.factor(year(Begin.Date)),
                  Total.Damage = (Property.Damage + Crop.Damage) * 1e-9)
hurricane.agg <- ddply(hurricane.data, ~ Year, summarize,
                       Total.Damage=sum(Total.Damage))
ggplot(hurricane.data, aes(Year, Total.Damage)) +
    geom_bar(stat="identity") + 
    labs(y="Crop and Property Damage [Billions of $]",
         title="Total Hurricane Damage by Year")
```