---
title: "Simulations of an exponential distribution"
output: pdf_document
widgets: mathjax
---

The idea here is to use some basic simulations of the exponential
distribution to explore the convergence of the mean and the
applicability of the central limit theorem for confidence interval
estimation using an exponential distribution with parameter $\lambda=0.2$.

## Basic Properties

Now we want to simulate a large number of draws from this distribution.
Specifically, we will simulate drawing 40 values and taking the mean
each time.  The expectation value of the mean is $E[x] = 1 / \lambda$,
and of the variance of those means is $V[x] = 1 / \lambda^2 n$ (where n is
40 in this case).

First, do the actual sims by defining a function which returns
the mean values in a frame:
```{r basicsim, echo=FALSE, warning=FALSE}
do.exp.sims <- function(lam, nsims, ndraws) {
    vals <- rexp(ndraws * nsims, rate=lam)
    dim(vals) <- c(nsims, ndraws)
    data.frame(means=apply(vals, 1, mean))}
set.seed(49898453)
nsims <- 2000
ndraws <- 40
lambda <- 0.2
sim.res <- do.exp.sims(lambda, nsims, ndraws)
```

Now compare the mean values to the theoretical expectation, including
the normal appoximation.
```{r meanplot, fig.width=5, fig.height=3, echo=FALSE, warning=FALSE}
library(ggplot2)
dnorm.args <- list(mean=mean(sim.res$means), sd=sd(sim.res$means))
ggplot(sim.res, aes(means)) + 
    geom_histogram(fill="dark grey", binwidth=0.05, aes(y=..density..)) +
    labs(x="Mean of 40 sims", title="Exponential distribution") +
    geom_vline(xintercept=1/lambda, color="red", linetype="longdash", size=2) +
    geom_vline(xintercept=mean(sim.res$means), color="black") +
    stat_function(fun=dnorm, arg=dnorm.args, color="green")
```

The black line is the observed mean (`r format(mean(sim.res$means), 
digits=4)`), and the red dashed line is the expected mean 
(`r 1 / lambda`).  The green curve is the
corresponding normal distribution with the observed sample parameters.
The distribution is reasonably well described by a normal distribution --
in other words, our sample is large enough that the central limit theorem
is a decent approximation.

The measured variance of those means and the expected value are
given by:
```{r vars}
sample.variance = var(sim.res$means)
expected.variance = 1 / (lambda^2 * ndraws)
```
which have the values `r format(sample.variance, digits=4)` and 
`r format(expected.variance, digits=4)`, respectively, which are
reasonably close.

## Coverage

Now lets check the coverage against the exact and asymptotic calculation
for the 95% frequentist confidence interval.  The approximate 
asymptotic normal interval, and the fraction of sims that lie within it, are
```{r approxinterval}
sd.approx <- sqrt(sample.variance)
sample.mean <- mean(sim.res$means)
lim.plus <- sample.mean + 1.96 * sd.approx
lim.minus <- sample.mean - 1.96 * sd.approx
frac.covered <- mean(sim.res$means <= lim.plus & sim.res$means >= lim.minus)
```
which is `r paste(format(100*frac.covered,digits=3), "%", sep="")` compared 
with the desired coverage of 95%.  Not bad!

We can do this for a range of values of the exponential parameter.
Note that we use the sample parameters instead of the theoretical ones.
```{r covtest, fig.width=4, fig.height=3}
exp.cover <- function(lam, nsims, ndraws) {
    evals <- do.exp.sims(lam, nsims, ndraws)
    smn <- mean(evals$means)
    ssd <- sd(evals$means)
    lim.low <- smn - 1.96 * ssd
    lim.hi <- smn + 1.96 * ssd
    mean(evals$means >= lim.low & evals$means <= lim.hi)
}
lamvals <- exp(seq(log(0.02), log(10.0), length.out=20))
cover <- sapply(lamvals, exp.cover, 1000, ndraws)
ggplot(data.frame(lamvals, cover), aes(lamvals, cover)) + geom_line(size=2) +
    geom_hline(yintercept = 0.95, color="red", linetype="dashed") +
    labs(x=expression(lambda), y="Coverage") + ylim(0.7, 1) + 
    scale_x_log10()
```

The thick line is the measured coverage, the thinner dashed red line
the target value (95%).  So it works quite well!