---
title: "Tooth Growth Exploration"
output: pdf_document
---

## Basic Description of the Data

The data in question explore the effects of vitamin C on
tooth growth in hamsters as a function of dose (3 levels) and delivery
method (orange juice or ascorbic acid, which are labeled OJ and VC
respectively).  There are 10 measurements
per combination (for a total of 60).  I will treat these measurements
as independent (unpaired).  Start with a rough look at the data
using boxplots:
```{r data, fig.width=6, fig.height=3}
library(datasets)
library(ggplot2)
data(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
ggplot(ToothGrowth, aes(dose, len)) + 
    geom_boxplot(aes(fill=dose), binwidth=4) +
    facet_grid(. ~ supp)
```

It certainly appears that increases in dose increase the tooth length
within each delivery method.  It isn't clear, however, if the delivery
method makes much difference.

## Efficacy of different delivery methods

For the first question I use a student's T test assuming equal variance
between the two (one could test this assumption with an F-test (var.test 
in R), but this wasn't discussed in the lectures).  The sample standard
deviations of each population are 
`r format(sd(ToothGrowth[ToothGrowth$supp == "OJ", 1]), digits=3)`
(OJ) and `r format(sd(ToothGrowth[ToothGrowth$supp == "VC", 1]), digits=3)`
(VC) respectively, which at least appear plausibly close.
 
I want the two sided test, since we only care that they differ here.
```{r methodtesteq}
p.delivery.eq <- t.test(len ~ supp, ToothGrowth, var.equal=TRUE)[["p.value"]]
```

The resulting p-value is `r format(p.delivery.eq, digits=2)`, which
provides moderate evidence that they differ, but not quite up to the
95\% confidence level.  Hence, I do not claim that there is significant
evidence that the different delivery methods produce differing levels
of tooth growth.  

Note that this conclusion is not affected much
if we do not assume equal variances:
```{r methodtestneq}
p.delivery.neq <- t.test(len ~ supp, ToothGrowth, var.equal=FALSE)[["p.value"]]
```

which gives a p-value `r format(p.delivery.neq, digits=2)`.

## Efficacy of differing doses

Using the methods discussed in class, there is no easy way to compare
the three different dose levels.  One could imagine fitting a linear
model vs. dose, for example.  One way
to proceed is to use the t-test to compare pairs of dose values
(e.g., 0.5 vs. 2, 0.5 vs 1), and then correct for multiple tests.
A permutation approach could also be considered.
```{r dosedata}
dose.data_0p5_1p0 <- ToothGrowth[ToothGrowth$dose %in% c("0.5", "1"),]
dose.data_0p5_2p0 <- ToothGrowth[ToothGrowth$dose %in% c("0.5", "2"),]
dose.data_1p0_2p0 <- ToothGrowth[ToothGrowth$dose %in% c("1", "2"),]
```
Proceeding to the t-test, for simplicity I will not assume equal
variances here.  It seems reasonable that the experimenters were 
testing whether increased vitamin C leads to increased tooth growth, 
so a one sided test is appropriate.
```{r dosettest}
p.dose.neq_0p5_1p0 <- t.test(len ~ dose, dose.data_0p5_1p0, 
                            alternative="less", var.equal=FALSE)[["p.value"]]
p.dose.neq_0p5_2p0 <- t.test(len ~ dose, dose.data_0p5_2p0, 
                            alternative="less", var.equal=FALSE)[["p.value"]]
p.dose.neq_1p0_2p0 <- t.test(len ~ dose, dose.data_1p0_2p0, 
                            alternative="less", var.equal=FALSE)[["p.value"]]
```

These p-values are extremely small (`r format(p.dose.neq_0p5_1p0)`,
`r format(p.dose.neq_0p5_2p0)`, and `r format(p.dose.neq_1p0_2p0)`.
respectively).  We could correct them using `p.adjust` for the
fact that we did multiple tests (e.g., by using the Bonferroni
correction); however, in this case that isn't useful because the
number of tests we did is quite modest, and the values already
pass any reasonable requirement.  Such small p-values are not
really reliable at that level anyways because the various assumptions
underlying our test are not really true at that level for any real data,
and so correcting them for multiple tests is not warranted.
A similar comment applies to the question of whether one or two-sided
tests are appropriate.

## Conclusions

* There is not significant evidence for any relation between
 delivery method and tooth growth.
* In contrast, there is very strong statistical evidence for
 a positive relation between dose level and tooth growth
 when different delivery methods are pooled.
 
Some of the assumptions include: 

* The underlying populations are approximately normally distributed
  within each group.
* That confounding factors have been controlled in some fashion,
  such as via random assignment of hamsters to groups.
* The observations are not paired (so there were a total of 60 
  hamsters involved).
* The results do not depend on assumptions of equal variance.